### 2020-11-22 manuscript/manuscript.rmd
suceefully rerun the manuscript and update Fig2 taxon name
updated manuscript format

---
bibliography: manuscript.bib
csl: nature.csl
header-includes:
   - \usepackage{lineno}
   - \linenumbers
output:
  html_document: default
  pdf_document:
    latex_engine: xelatex
  word_document: default
  github_document: default
---

@@

## Summary

@@

Historically, there was little debate about the root of the animal tree of life. Porifera-sister (Fig. 1E), the hypothesis that the animal root marks the divergence of Porifera (sponges) from all other animals (Fig. 1B), was widely accepted though rarely tested. By contrast, there has long been uncertainty about the placement of Ctenophora (comb jellies) (Fig. 1D) in the animal tree of life [@Wallberg:2004ws]. The first phylogenomic study to include ctenophores [@Dunn:2008ky] suggested a new hypothesis, now referred to as Ctenophora-sister, that ctenophores rather than sponges are our most distant living animal relative (Fig. 1A). Since then, many more phylogenomic studies have been published (Fig. 2), with some analyses finding support for Ctenophora-sister, some for Porifera-sister, and some neither [@King:2017ie] (Extended Data Table 1, Supplementary Information section S1). The extensive technical variation across these studies has been important to advancing our understanding of the sensitivity of these analyses, demonstrating for example that outgroup and model selection can have a large impact on these results (Fig. 3A). But the extensive technical variation has also made it difficult to synthesize these results to understand the underlying causes of this sensitivity. Several factors make resolution of the root of the animal tree a particularly challenging problem. For one, the nodes in question are the deepest in the animal tree of life. Another factor that has been invoked is branch lengths (*e.g.* Extended Data Figs. 1, 2), which are impacted by both divergence times and shifts in rate of evolution. Some sponges have a longer root to tip length, indicating an accelerated rate of evolution in those lineages. The stem branch of Ctenophora is longer than Porifera stem which, together with a more typical root-to-tip distance, is consistent with a more recent radiation of extant ctenophores [@podar2001] than extant poriferans. The longer ctenophore stem has led some to suggest that Ctenophora-sister could be an artifact of long-branch attraction to outgroups [@pisani2015genomic].

To advance this debate it is critical to examine variation in methods and results in a standardized and systematic framework. Here, we synthesize data and results from 15 previous phylogenomic studies that tested the Ctenophora-sister and Porifera-sister hypotheses (Fig. 2, Extended Data Table 1). This set includes all phylogenomic studies of amino acid sequence data published before 2019 for which we could obtain data matrices with gene partition annotations. Among these 15 studies, three studies [@pisani2015genomic; @whelan2016let; @feuda2017improved] are based entirely on previously published data, and gene-partition data is not available from one study [@Pick:2010eb].

@@

High-throughput sequencing allows investigators to readily assemble matrices with hundreds or thousands of protein-coding genes from a broad diversity of animal species (Extended Data Table 1). Studies of animal phylogeny have used a wide variety of different approaches to identifying and selecting genes and taxa for their matrices. As a result, the particular genes selected for analysis vary widely (Fig. 2, Extended Data Fig. 3). Gene sampling varies in several ways, including in the fractions of single-copy orthologs (*e.g.* BUSCO genes) and ribosomal protein genes in the matrix (Extended Data Fig. 4).

@@

Ingroup taxon sampling also varies widely between studies (Fig. 2, Extended Data Fig. 3). Sampling of ingroup taxa (animals) in early studies was biased toward Bilateria. Sampling of non-bilaterian animals, including sponges and ctenophores, has improved over time (Extended Data Fig. 3). Within each clade, there is often considerable variation in taxon sampling and therefore often little species overlap across studies (Extended Data Fig. 3). This variation is in part because newer sequencing technologies in more recent studies are usually not applied to the exact same species that were included in earlier studies.

@@

![](figures/Figure3_published.png)
**Fig. 3.**  Summary of phylogenomic results from previous studies and reanalysis conducted in this study. Each point represents a phylogenetic analysis with a support of bootstrap values less than `r bootstrap_threshold`% or a posterior probability less than `r posterior_prob_threshold`% are considered as unresolved. The clades are organized with increased outgroup sampling higher on the vertical axis, and the models are organized in general by increasing complexity in terms of numbers of parameters to the right on the horizontal axis. (A) Analyses transcribed from the literature (related to Supplementary Table 1). (B) New phylogenomic analyses conducted in this study (related to Supplementary Table 2). Over 1,011,765 CPU hours (~115 CPU years) were used for this study.

We parsed `r analyses_processed %>% filter(status=="published") %>% nrow()` previous phylogenetic analyses from `r analyses_processed %>% filter(status=="published") %>% pull (manuscript) %>% unique() %>% length()` studies (Fig. 3A and Supplementary Table 1). The conclusions of five studies strongly favor Porifera-sister and ten favor Ctenophora-sister (Extended Data Table 1). Three studies are based entirely on previously published data, and the remainder add data for one or more species.s

@@

One of the challenges of interpreting support for the placement of the animal root across published studies is that different programs, software versions, and settings have been used across analyses. This extensive variation makes it difficult to identify the primary factors that lead to different results. Here we first reanalyze the primary matrices from each study under the same conditions with IQ-TREE [@nguyen2014iq] with multiple evolutionary models. We selected this tool because it has greater model flexibility than most other phylogenetic tools [@zhou2017evaluating] (Fig. 3B; Supplementary Table 3). Importantly, it has C models (C10-C60 equilibrium frequencies) [@si2008empirical] that, like CAT, allow models to accommodate heterogeneity in equilibrium amino acid frequencies across sites.

@@

For all but the three largest matrices, we tested the relative fit of a variety of models, both with and without C10-C60 accommodation of site heterogeneity in equilibrium frequencies, using ModelFinder [@kalyaanamoorthy2017modelfinder] in IQ-TREE. In all cases, models with C60 fit these matrices better than the site-homogeneous models. This is consistent with the importance of accommodating site heterogeneity noted by previous investigators [@Lartillot:2004dq; @Philippe:2009hh; @nosenko2013deep; @pisani2015genomic; @simion2017large; @feuda2017improved]. We then inferred support under the best-fit model (Supplementary Table 3), except for the three largest matrices where we used LG+C60. We then analyzed each matrix under a panel of standard site-heterogeneous and site-homogeneous models, including WAG, GTR and Poisson+C60 (Supplementary Table 3).

All IQ-TREE analyses, apart from unresolved analyses (for Moroz2014_3d and all Nosenko2013 matrices), supported Ctenophora-sister (Fig. 3B). No IQ-TREE analyses supported Porifera-sister, including those that restrict outgroup sampling to Choanoflagellatea and use models with site-heterogeneous equilibrium frequencies (Fig. 3B lower right; Supplementary Table 2), the conditions under which published PhyloBayes CAT analyses recover strong support for Porifera-sister (Fig. 3A). To further verify this difference in a controlled manner we reran PhyloBayes analyses with CAT, using both Poisson and GTR substitution matrices, for some matrices that had led to support for Porifera-sister in published analyses. Consistent with published results, some of these supported Porifera-sister.

@@

A primary difference between the C (like C60) and CAT site-heterogeneous models is the number of equilibrium frequency categories. The standard CAT model employs a Dirichlet process prior to inferring the number of equilibrium frequency categories, so the number of categories is variable [@Lartillot:2004dq]. IQ-TREE implements C models [@si2008empirical] with a fixed number of categories, from 10 (C10) to 60 (C60). Differences in analysis results could therefore be due to differences in the number of categories. The number of categories inferred by CAT in PhyloBayes can be very high (Supplementary Table 4), with a mean here of `r cat_categories$n_categories_poisson %>% mean() %>% round(1)` categories for Poisson+CAT analyses in all matrices and `r cat_categories$n_categories_gtr  %>% mean(na.rm=TRUE) %>% round(1)` categories for GTR+CAT analyses in several representative matrices. This requires a very large number of additional estiamted parameters.

We examined the specific impact of this large difference in category number on animal root position. It is currently not possible to use more than 60 categories for C models in IQ-TREE, but the number of categories can be set *a priori* in Phylobayes using nCAT. We therefore varied the number of categories in Phylobayes analyses (Fig. 4; Supplementary Table 5). We found that Poisson+nCAT60 analyses in PhyloBayes, like Poisson+C60 and WAG+C60 IQ-TREE analyses, provide strong support of Ctenophora-sister. This indicates that the difference in results between unconstrained CAT analyses in PhyloBayes and C60 analyses in IQ-TREE is not due to differences in the software or other implementation factors, but due to the large difference, in excess of ten-fold, in the number of site categories. When we increased the number of categories in PhyloBayes nCAT analyses, we observe the transition from support for Ctenophora-sister to an unresolved root to Porifera-sister (Fig. 4). For example, for the Whelan2017_strict matrix this transition occurs between 60 -- 120 categories when using Poisson+nCAT model. Due to computational limitations of GTR models, we only ran GTR+CAT and GTR+nCAT60 models on representative matrices with Choanozoa sampling. We found that for Whelan2017 matrices, support shifted from Porifera-sister with Poisson+CAT model to Ctenophora-sister using GTR+CAT model. Moreover, we also found all results strongly supported Ctenophora-sister with GTR+nCAT60 models.

These results further clarify when analyses support Porifera-sister (*e.g.* Whelan2017: Extended Data Fig. 1; Philippe2009: Extended Data Fig. 2): when outgroup sampling is restricted (Choanozoa), when a Poisson (rather than a GTR) exchange matrix is used, and when a very large number of site categories is used (unconstrained CAT, giving hundreds of equilibrium frequency categories). Analyses under other conditions either support Ctenophora-sister or are unresolved. This is consistent across published analyses and our new panels of analyses. The question, then, isn’t why similar analyses give different results, but how we should interpret variation in results when we run analyses that differ in these specific respects. If we fix the first two features to conditions that are necessary for Porifera-sister support (Choanozoa taxon sampling and a Poisson exchange matrix), there are several insights that we can glean from examining how the number of equilibrium frequency categories impacts results that sheds light on interpretation of those results.

The first questions regard model fit. ModelFinder selects site heterogeneous C60 models according to BIC (Supplementary Table 3), but IQ-TREE often gives a warning under C60 that the model may overfit, with too many categories, because some mixture weights are close to 0 (Supplementary Table 6). In PhylobBayes, Cross-validation is a reliable and suggested approach to model evaluation [@lartillot2009phylobayes]. We elvaluated  Poisson+CAT and Poisson+nCAT60 models with cross-validation in PhyloBayes for the Whelan2017_strict and Philippe2009_Choanozoa matrices. For both matrices, we found nearly identical distributions of cross-validation scores for Poisson+nCAT60 and Poisson+CAT models (Extended Data Fig. 5). A t-test analysis shows that there is no significant difference between them for either matrix (Whelan2017_strict: $p-value=`r P_value_Whelan2017_strict`$; Philippe2009_Choanozoa: $p-value=`r P_value_Philippe2009`$). Cross-validation therefore does not indicate that the unconstrained CAT models are a better fit than the models with 60 categories.  This suggests there is not a reason to favor the narrow analysis conditions (Choanozoa taxon sampling, Poisson exchange matrix, and unconstrained CAT models that use hundreds of categories) under which Porifera-sister are obtained.

In a site-heterogeneous model, each site is allocated to one of the categories in each sample. Different categories can have very different numbers of sites. We examine category allocations in the last chain samples of analyses of the Philippe2009_Choanozoa and Whelan2017_strict matrices (Extended Data Fig. 6A-B), which have `r n_categories_phil_cat` and `r n_categories_whel_cat` categories, respectively. The fraction of sites allocated to the 50% of the least frequent categories is `r mid_phil_cat`% in the analysis of the Philippe2009_Choanozoa matrix and `r mid_whel_cat`% in the analysis of the Whelan2017_strict matrix. This long tail of categories that apply to a very small fraction of sites is in stark contrast to nCAT60 analyses, constrained to 60 categories, which have no such long tail of rare categories (Extended Data Fig. 6A-B). The fact that so many categories apply to such a small fraction of data may help explain why increasing the number of categories more than ten-fold has so little impact on the predictive power of these far more complex models.

@@

```
![](figures/Figure4_manuscript.png)
**Fig. 4.**  Sensitivity analyses with representative data matrices were analyzed by different number of equilibrium frequency categories (nCAT) in PhyloBayes. Statistical support values (posterior probabilities) were obtained from three data matrices using the site-heterogeneous Poisson+CAT model with different categories. (A) Phlippe2009_Choanozoa; (B) Philippe2009_Holozoa; (C) Whelan2017_full_Choanozoa; (D) Whelan2017_strict. Statistical support for Ctenophora-sister and Porifera-sister is indicated in orange and green, respectively. Support values from the sensitivity analyses are shown in Supplementary Table 5.

@@

To further explore the phylogenetic signal of different models we discussed above, we quantified the phylogenetic signal for Porifera-sister and Ctenophora-sister topologies across three representative data matrices when varying outgroup sampling and model (Extended Data Fig. 7). By calculating differences in log-likelihood scores for these topologies for every gene ($\Delta$|lnL|) in each matrix when using site-homogeneous models in IQ-TREE, we found that the Ctenophora-sister had the higher proportions of supporting genes in every analysis. Moreover, outgroup choice has little impact on the distribution of the support for phylogenetic signals in analyses with site-homogeneous models. This finding is largely consistent with the previously observed distribution of support for Ctenophora-sister in other data matrices [@shen2017contentious].

Although a higher proportion of genes support Ctenophora-sister with site-heterogeneous C60 models, the phylogenetic signal decreases in many genes using C60 models compared to site-homogeneous models. In an extreme case, in matrices from Ryan2013_est nearly 30% of genes changed from strong $\Delta$|lnL|>2) to weak Ctenophora-sister signal ($\Delta$|lnL|<2) (Extended Data Fig. 7; Supplementary Table 7). In contrast to the C60 models, there is a major increase in phylogenetic signal in Poisson+CAT models in PhyloBayes towards Porifera-sister, and outgroup choice has a major effect of the distribution of phylogenetic signal (Extended Data Fig. 7). For example, in Whelan2017_full matrix, we found that the number of genes that support Ctenophora-sister in analyses with CAT decreases from 57.5% in matrices with distant outgroups (Holozoa) to 35.4% when outgroups are restricted (Choanozoa, Supplementary Table 7).

@@

Other approaches have been taken to addressing base compositional heterogeneity across taxa. For example, Feuda *et al.* [@Feuda:2017ew] recoded the full set of twenty amino acids into six groups of amino acids. These groups tend to have more frequent evolutionary changes within them than between them [@Susko:2007ds]. Recoding could, like CAT and C models, address variation across sites, but it could also accommodate variation across lineages, and it was suggested that this approach favors Porifera-sister [@Feuda:2017ew]. However, our own analyses (Supplementary Information section S3, Supplementary Figs. 8-9) comparing the performance of random and empirical recoding indicate that the impact of data recoding is largely due to discarding information, not accommodating variation in amino acid composition. Consistent with a recent simulation study on data recoding [@hernandez2019six], these findings indicate that recoding can be a problematic method for addressing heterogeneity.

@@

## Methods

@@

We retrieved matrices from each publication (Extended Data Table 1), storing the raw data in this manuscript's version control repository. We manually made some formatting changes to make the batch processing of the matrices work well, *e.g.* standardizing the format of Nexus CHARSET blocks. All changes made are tracked with git.

@@

#### Phylogenetic analyses in IQ-TREE
To investigate the phylogenetic hypotheses and distribution of phylogenetic signal in studies aiming to find the root position of animal phylogeny, we considered 16 data matrices from all phylogenomic studies that were constructed from EST, transcriptomic, or genomic data (Extended Data Table 1). Because different choices of substitution models could largely influence phylogenetic inference of the placement of the root position of animal phylogeny (*e.g.* site-heterogeneous vs. site-homogeneous models), we first investigated model-fit from each matrix using ModelFinder in IQ-TREE v1.6.7, including site-heterogenous C10 to C60 profile mixture models (C60 models) as variants of the CAT models in ML framework (C10-C60 model were included for model comparison via -madd option). We included models that are commonly used in previous analyses, including site-homogeneous Poisson, WAG, LG, GTR models plus C10-C60 models in the model testing. For computational efficiency, the GTR+C60 models were not included in model testing since it requires to estimate over 10,000 parameters. For large matrices like those from Hejnol2009, Borrowiec2015, and Simion2017, model testing is also not computational feasible so only LG+C60 models were used since LG/WAG+C60 models were suggested as the best-fit model in all other matrices.

@@

The 20 state matrices are the same across all analyses since they are not recoded. Since all 20 state matrices are the same, variation between 20-state results (as in the left side of each pane of Extended Data Fig. 9) gives insight into the technical variance of the inference process.

@@

The statistics presented in  Extended Data Fig. 8A were parsed from the Feuda *et al.* manuscript into the file `tidy_table_3.tsv` and rendered for display with the code in `manuscript.rmd`.

@@

We thank members of the Dunn and Rokas laboratories for discussions and comments. We are grateful to Steve Haddock, Kyle David, and Nathan Whelan for feedback on an earlier version of this manuscript. We are grateful to Nicole King, Daniel Richter, Paul Lewis, Warren Francis and Joseph Ryan for feedback on the preprint version of this manuscript. We thank the Yale Center for Research Computing for use of the research computing infrastructure, specifically the Farnam HPC cluster. Yuanning Li was partially supported by a scholarship from the China Scholarship Council (CSC) for studying and living abroad. This work was conducted in part using the resources of the Advanced Computing Center for Research and Education (ACCRE) at Vanderbilt University. Research in A.R.'s lab is supported by the National Science Foundation (DEB-1442113), the National Institutes of Health / National Institute of Allergy and Infectious Diseases (1R56AI146096-01A1), the Guggenheim Foundation, and the Burroughs Wellcome Fund.

@@

## Extended Tables

**Extended Data Table 1.** Overview of data matrices used in this study.
```{r table_study_summary}
table_study_summary = read_tsv("../data_processed/tables/study_summary.tsv")
opts <- options(knitr.kable.NA = "")
knitr::kable(table_study_summary, digits = 3, format.args = list(big.mark = ",",
  scientific = FALSE)) %>%
  kable_styling(latex_options="scale_down")

@@

## Extended Figures

@@

**Extended Data Fig. 3.** Pairwise overlap between each of the primary matrices considered here (related to Supplementary Table 8). Horizontal size is proportional to the number of genes sampled, vertical size to the number of taxa sampled. The horizontal intersection shows the proportions of shared genes, the vertical intersection shows the proportions of shared taxa.

>>>>

---
bibliography: manuscript.bib
csl: nature.csl
header-includes:
   - \usepackage{lineno}
   - \linenumbers
output:
  pdf_document:
    latex_engine: xelatex
  html_document: default
  word_document: default
  github_document: default
---

@@

## Abstract

@@

Historically, there was little debate about the root of the animal tree of life. Porifera-sister (Fig. 1E), the hypothesis that the animal root marks the divergence of Porifera (sponges) from all other animals (Fig. 1B), was widely accepted though rarely tested. By contrast, there has long been uncertainty about the placement of Ctenophora (comb jellies) (Fig. 1D) in the animal tree of life [@Wallberg:2004ws]. The first phylogenomic study to include ctenophores [@Dunn:2008ky] suggested a new hypothesis, now referred to as Ctenophora-sister, that ctenophores rather than sponges are our most distant living animal relative (Fig. 1A). Since then, many more phylogenomic studies have been published (Fig. 2), with some analyses finding support for Ctenophora-sister, some for Porifera-sister, and some neither [@King:2017ie] (Table 1, Supplementary Information section S1). The extensive technical variation across these studies has been important to advancing our understanding of the sensitivity of these analyses, demonstrating for example that outgroup and model selection can have a large impact on these results (Fig. 3A). But the extensive technical variation has also made it difficult to synthesize these results to understand the underlying causes of this sensitivity. Several factors make resolution of the root of the animal tree a particularly challenging problem. For one, the nodes in question are the deepest in the animal tree of life. Another factor that has been invoked is branch lengths (*e.g.* Figs. S1, S2), which are impacted by both divergence times and shifts in rate of evolution. Some sponges have a longer root to tip length, indicating an accelerated rate of evolution in those lineages. The stem branch of Ctenophora is longer than Porifera stem which, together with a more typical root-to-tip distance, is consistent with a more recent radiation of extant ctenophores [@podar2001] than extant poriferans. The longer ctenophore stem has led some to suggest that Ctenophora-sister could be an artifact of long-branch attraction to outgroups [@pisani2015genomic].

To advance this debate it is critical to examine variation in methods and results in a standardized and systematic framework. Here, we synthesize data and results from 15 previous phylogenomic studies that tested the Ctenophora-sister and Porifera-sister hypotheses (Fig. 2, Table 1). This set includes all phylogenomic studies of amino acid sequence data published before 2019 for which we could obtain data matrices with gene partition annotations. Among these 15 studies, three studies [@pisani2015genomic; @whelan2016let; @feuda2017improved] are based entirely on previously published data, and gene-partition data is not available from one study [@Pick:2010eb].

@@

High-throughput sequencing allows investigators to readily assemble matrices with hundreds or thousands of protein-coding genes from a broad diversity of animal species (Table 1). Studies of animal phylogeny have used a wide variety of different approaches to identifying and selecting genes and taxa for their matrices. As a result, the particular genes selected for analysis vary widely (Fig. 2, Fig. S3). Gene sampling varies in several ways, including in the fractions of single-copy orthologs (*e.g.* BUSCO genes) and ribosomal protein genes in the matrix (Fig. S4).

@@

Ingroup taxon sampling also varies widely between studies (Fig. 2, Fig. S3). Sampling of ingroup taxa (animals) in early studies was biased toward Bilateria. Sampling of non-bilaterian animals, including sponges and ctenophores, has improved over time (Fig. S3). Within each clade, there is often considerable variation in taxon sampling and therefore often little species overlap across studies (Fig. S3). This variation is in part because newer sequencing technologies in more recent studies are usually not applied to the exact same species that were included in earlier studies.

@@

![](figures/Figure3_published.png)
**Fig. 3.**  Summary of phylogenomic results from previous studies and reanalysis conducted in this study. Each point represents a phylogenetic analysis with a support of bootstrap values less than `r bootstrap_threshold`% or a posterior probability less than `r posterior_prob_threshold`% are considered as unresolved. The clades are organized with increased outgroup sampling higher on the vertical axis, and the models are organized in general by increasing complexity in terms of numbers of parameters to the right on the horizontal axis. (A) Analyses transcribed from the literature (related to Table S1). (B) New phylogenomic analyses conducted in this study (related to Table S2). Over 1,011,765 CPU hours (~115 CPU years) were used for this study.

We parsed `r analyses_processed %>% filter(status=="published") %>% nrow()` previous phylogenetic analyses from `r analyses_processed %>% filter(status=="published") %>% pull (manuscript) %>% unique() %>% length()` studies (Fig. 3A and Table S1). The conclusions of five studies strongly favor Porifera-sister and ten favor Ctenophora-sister (Table 1). Three studies are based entirely on previously published data, and the remainder add data for one or more species.s

@@

One of the challenges of interpreting support for the placement of the animal root across published studies is that different programs, software versions, and settings have been used across analyses. This extensive variation makes it difficult to identify the primary factors that lead to different results. Here we first reanalyze the primary matrices from each study under the same conditions with IQ-TREE [@nguyen2014iq] with multiple evolutionary models. We selected this tool because it has greater model flexibility than most other phylogenetic tools [@zhou2017evaluating] (Fig. 3B; Table S3). Importantly, it has C models (C10-C60 equilibrium frequencies) [@si2008empirical] that, like CAT, allow models to accommodate heterogeneity in equilibrium amino acid frequencies across sites.

@@

For all but the three largest matrices, we tested the relative fit of a variety of models, both with and without C10-C60 accommodation of site heterogeneity in equilibrium frequencies, using ModelFinder [@kalyaanamoorthy2017modelfinder] in IQ-TREE. In all cases, models with C60 fit these matrices better than the site-homogeneous models. This is consistent with the importance of accommodating site heterogeneity noted by previous investigators [@Lartillot:2004dq; @Philippe:2009hh; @nosenko2013deep; @pisani2015genomic; @simion2017large; @feuda2017improved]. We then inferred support under the best-fit model (Table S3), except for the three largest matrices where we used LG+C60. We then analyzed each matrix under a panel of standard site-heterogeneous and site-homogeneous models, including WAG, GTR and Poisson+C60 (Table S3).

All IQ-TREE analyses, apart from unresolved analyses (for Moroz2014_3d and all Nosenko2013 matrices), supported Ctenophora-sister (Fig. 3B). No IQ-TREE analyses supported Porifera-sister, including those that restrict outgroup sampling to Choanoflagellatea and use models with site-heterogeneous equilibrium frequencies (Fig. 3B lower right; Table S2), the conditions under which published PhyloBayes CAT analyses recover strong support for Porifera-sister (Fig. 3A). To further verify this difference in a controlled manner we reran PhyloBayes analyses with CAT, using both Poisson and GTR substitution matrices, for some matrices that had led to support for Porifera-sister in published analyses. Consistent with published results, some of these supported Porifera-sister.

@@

A primary difference between the C (like C60) and CAT site-heterogeneous models is the number of equilibrium frequency categories. The standard CAT model employs a Dirichlet process prior to inferring the number of equilibrium frequency categories, so the number of categories is variable [@Lartillot:2004dq]. IQ-TREE implements C models [@si2008empirical] with a fixed number of categories, from 10 (C10) to 60 (C60). Differences in analysis results could therefore be due to differences in the number of categories. The number of categories inferred by CAT in PhyloBayes can be very high (Table S4), with a mean here of `r cat_categories$n_categories_poisson %>% mean() %>% round(1)` categories for Poisson+CAT analyses in all matrices and `r cat_categories$n_categories_gtr  %>% mean(na.rm=TRUE) %>% round(1)` categories for GTR+CAT analyses in several representative matrices. This requires a very large number of additional estiamted parameters.

We examined the specific impact of this large difference in category number on animal root position. It is currently not possible to use more than 60 categories for C models in IQ-TREE, but the number of categories can be set *a priori* in Phylobayes using nCAT. We therefore varied the number of categories in Phylobayes analyses (Fig. 4; Table S5). We found that Poisson+nCAT60 analyses in PhyloBayes, like Poisson+C60 and WAG+C60 IQ-TREE analyses, provide strong support of Ctenophora-sister. This indicates that the difference in results between unconstrained CAT analyses in PhyloBayes and C60 analyses in IQ-TREE is not due to differences in the software or other implementation factors, but due to the large difference, in excess of ten-fold, in the number of site categories. When we increased the number of categories in PhyloBayes nCAT analyses, we observe the transition from support for Ctenophora-sister to an unresolved root to Porifera-sister (Fig. 4). For example, for the Whelan2017_strict matrix this transition occurs between 60 -- 120 categories when using Poisson+nCAT model. Due to computational limitations of GTR models, we only ran GTR+CAT and GTR+nCAT60 models on representative matrices with Choanozoa sampling. We found that for Whelan2017 matrices, support shifted from Porifera-sister with Poisson+CAT model to Ctenophora-sister using GTR+CAT model. Moreover, we also found all results strongly supported Ctenophora-sister with GTR+nCAT60 models.

These results further clarify when analyses support Porifera-sister (*e.g.* Whelan2017: Fig. S1; Philippe2009: Fig. S2): when outgroup sampling is restricted (Choanozoa), when a Poisson (rather than a GTR) exchange matrix is used, and when a very large number of site categories is used (unconstrained CAT, giving hundreds of equilibrium frequency categories). Analyses under other conditions either support Ctenophora-sister or are unresolved. This is consistent across published analyses and our new panels of analyses. The question, then, isn’t why similar analyses give different results, but how we should interpret variation in results when we run analyses that differ in these specific respects. If we fix the first two features to conditions that are necessary for Porifera-sister support (Choanozoa taxon sampling and a Poisson exchange matrix), there are several insights that we can glean from examining how the number of equilibrium frequency categories impacts results that sheds light on interpretation of those results.

The first questions regard model fit. ModelFinder selects site heterogeneous C60 models according to BIC (Table S3), but IQ-TREE often gives a warning under C60 that the model may overfit, with too many categories, because some mixture weights are close to 0 (Table S6). In PhylobBayes, Cross-validation is a reliable and suggested approach to model evaluation [@lartillot2009phylobayes]. We elvaluated  Poisson+CAT and Poisson+nCAT60 models with cross-validation in PhyloBayes for the Whelan2017_strict and Philippe2009_Choanozoa matrices. For both matrices, we found nearly identical distributions of cross-validation scores for Poisson+nCAT60 and Poisson+CAT models (Fig. S5). A t-test analysis shows that there is no significant difference between them for either matrix (Whelan2017_strict: $p-value=`r P_value_Whelan2017_strict`$; Philippe2009_Choanozoa: $p-value=`r P_value_Philippe2009`$). Cross-validation therefore does not indicate that the unconstrained CAT models are a better fit than the models with 60 categories.  This suggests there is not a reason to favor the narrow analysis conditions (Choanozoa taxon sampling, Poisson exchange matrix, and unconstrained CAT models that use hundreds of categories) under which Porifera-sister are obtained.

In a site-heterogeneous model, each site is allocated to one of the categories in each sample. Different categories can have very different numbers of sites. We examine category allocations in the last chain samples of analyses of the Philippe2009_Choanozoa and Whelan2017_strict matrices (Fig. S 6A-B), which have `r n_categories_phil_cat` and `r n_categories_whel_cat` categories, respectively. The fraction of sites allocated to the 50% of the least frequent categories is `r mid_phil_cat`% in the analysis of the Philippe2009_Choanozoa matrix and `r mid_whel_cat`% in the analysis of the Whelan2017_strict matrix. This long tail of categories that apply to a very small fraction of sites is in stark contrast to nCAT60 analyses, constrained to 60 categories, which have no such long tail of rare categories (Fig. S6A-B). The fact that so many categories apply to such a small fraction of data may help explain why increasing the number of categories more than ten-fold has so little impact on the predictive power of these far more complex models.

@@

```
![](figures/Figure4_manuscript.png)
**Fig. 4.**  Sensitivity analyses with representative data matrices were analyzed by different number of equilibrium frequency categories (nCAT) in PhyloBayes. Statistical support values (posterior probabilities) were obtained from three data matrices using the site-heterogeneous Poisson+CAT model with different categories. (A) Phlippe2009_Choanozoa; (B) Philippe2009_Holozoa; (C) Whelan2017_full_Choanozoa; (D) Whelan2017_strict. Statistical support for Ctenophora-sister and Porifera-sister is indicated in orange and green, respectively. Support values from the sensitivity analyses are shown in Table S5.

@@

To further explore the phylogenetic signal of different models we discussed above, we quantified the phylogenetic signal for Porifera-sister and Ctenophora-sister topologies across three representative data matrices when varying outgroup sampling and model (Fig. S7). By calculating differences in log-likelihood scores for these topologies for every gene ($\Delta$|lnL|) in each matrix when using site-homogeneous models in IQ-TREE, we found that the Ctenophora-sister had the higher proportions of supporting genes in every analysis. Moreover, outgroup choice has little impact on the distribution of the support for phylogenetic signals in analyses with site-homogeneous models. This finding is largely consistent with the previously observed distribution of support for Ctenophora-sister in other data matrices [@shen2017contentious].

Although a higher proportion of genes support Ctenophora-sister with site-heterogeneous C60 models, the phylogenetic signal decreases in many genes using C60 models compared to site-homogeneous models. In an extreme case, in matrices from Ryan2013_est nearly 30% of genes changed from strong $\Delta$|lnL|>2) to weak Ctenophora-sister signal ($\Delta$|lnL|<2) (Fig. S7; Table S7). In contrast to the C60 models, there is a major increase in phylogenetic signal in Poisson+CAT models in PhyloBayes towards Porifera-sister, and outgroup choice has a major effect of the distribution of phylogenetic signal (Fig. S7). For example, in Whelan2017_full matrix, we found that the number of genes that support Ctenophora-sister in analyses with CAT decreases from 57.5% in matrices with distant outgroups (Holozoa) to 35.4% when outgroups are restricted (Choanozoa, Table S7).

@@

Other approaches have been taken to addressing base compositional heterogeneity across taxa. For example, Feuda *et al.* [@Feuda:2017ew] recoded the full set of twenty amino acids into six groups of amino acids. These groups tend to have more frequent evolutionary changes within them than between them [@Susko:2007ds]. Recoding could, like CAT and C models, address variation across sites, but it could also accommodate variation across lineages, and it was suggested that this approach favors Porifera-sister [@Feuda:2017ew]. However, our own analyses (Supplementary Information section S3, Figs. S8-S9) comparing the performance of random and empirical recoding indicate that the impact of data recoding is largely due to discarding information, not accommodating variation in amino acid composition. Consistent with a recent simulation study on data recoding [@hernandez2019six], these findings indicate that recoding can be a problematic method for addressing heterogeneity.

@@

## Supplementary materials

## Materials and Methods

@@

We retrieved matrices from each publication (Table 1), storing the raw data in this manuscript's version control repository. We manually made some formatting changes to make the batch processing of the matrices work well, *e.g.* standardizing the format of Nexus CHARSET blocks. All changes made are tracked with git.

@@

#### Phylogenetic analyses in IQ-TREE
To investigate the phylogenetic hypotheses and distribution of phylogenetic signal in studies aiming to find the root position of animal phylogeny, we considered 16 data matrices from all phylogenomic studies that were constructed from EST, transcriptomic, or genomic data (Table 1). Because different choices of substitution models could largely influence phylogenetic inference of the placement of the root position of animal phylogeny (*e.g.* site-heterogeneous vs. site-homogeneous models), we first investigated model-fit from each matrix using ModelFinder in IQ-TREE v1.6.7, including site-heterogenous C10 to C60 profile mixture models (C60 models) as variants of the CAT models in ML framework (C10-C60 model were included for model comparison via -madd option). We included models that are commonly used in previous analyses, including site-homogeneous Poisson, WAG, LG, GTR models plus C10-C60 models in the model testing. For computational efficiency, the GTR+C60 models were not included in model testing since it requires to estimate over 10,000 parameters. For large matrices like those from Hejnol2009, Borrowiec2015, and Simion2017, model testing is also not computational feasible so only LG+C60 models were used since LG/WAG+C60 models were suggested as the best-fit model in all other matrices.

@@

The 20 state matrices are the same across all analyses since they are not recoded. Since all 20 state matrices are the same, variation between 20-state results (as in the left side of each pane of Fig. S9) gives insight into the technical variance of the inference process.

@@

The statistics presented in  Fig. S8A were parsed from the Feuda *et al.* manuscript into the file `tidy_table_3.tsv` and rendered for display with the code in `manuscript.rmd`.

@@

We thank members of the Dunn and Rokas laboratories for discussions and comments. We are grateful to Steve Haddock, Kyle David, Stephen Haddock and Nathan Whelan for feedback on an earlier version of this manuscript. We are grateful to Nicole King, Daniel Richter, Paul Lewis, Warren Francis and Joseph Ryan for feedback on the preprint version of this manuscript. We thank the Yale Center for Research Computing for use of the research computing infrastructure, specifically the Farnam HPC cluster. Yuanning Li was partially supported by a scholarship from the China Scholarship Council (CSC) for studying and living abroad. This work was conducted in part using the resources of the Advanced Computing Center for Research and Education (ACCRE) at Vanderbilt University. Research in A.R.'s lab is supported by the National Science Foundation (DEB-1442113), the National Institutes of Health / National Institute of Allergy and Infectious Diseases (1R56AI146096-01A1), the Guggenheim Foundation, and the Burroughs Wellcome Fund.

@@

## Tables

**Table 1.** Overview of data matrices used in this study.
```{r table_study_summary}
table_study_summary = read_tsv("../data_processed/tables/study_summary.tsv")
opts <- options(knitr.kable.NA = "")
knitr::kable(table_study_summary, digits = 3, format.args = list(big.mark = ",",
  scientific = FALSE)) %>%
  kable_styling(latex_options="scale_down")

@@

## Supplementary Figures

@@

**Fig. S3.** Pairwise overlap between each of the primary matrices considered here (related to Table S8). Horizontal size is proportional to the number of genes sampled, vertical size to the number of taxa sampled. The horizontal intersection shows the proportions of shared genes, the vertical intersection shows the proportions of shared taxa.
